---
title: "Science, Interrupted: How Scientists Respond to Funding Disruptions"
subtitle: "MTEI Seminar"  
author: "Wei Yang Tham<br><span style = 'font-size: 70%;'>Laboratory for Innovation Sciences at Harvard</span>"
date: "December 09, 2019"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      titleSlideClass: [left, middle]
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)

knitr::opts_chunk$set(warning = FALSE, message = FALSE, 
  comment = NA, dpi = 300, echo = FALSE, cache = F,
  fig.align = "center", out.width = "80%", fig.asp = 0.618)

```

```{r packages}

library(tidyverse)
library(magrittr)
library(here)
library(colorspace)
library(patchwork)

source(here("scripts/functions.R"))
```


```{r xaringan-themer, include=FALSE}
library(xaringanthemer)

duo(
  primary_color = "#fcedd1",
  secondary_color = "#2a488e",
  # secondary_color = "#4671d6",
  # text_color = "#2C3E50",
  header_font_google = google_font("Josefin Sans"),
  text_font_google   = google_font("Montserrat", "300", "300i"),
  code_font_google   = google_font("Droid Mono"),
  text_font_size = "25px", 
  text_slide_number_font_size = "0.0em"
)



```

```{r plots-setup}

theme_wyt <- function(base_size = 11, base_family = "", base_line_size = base_size/22, base_rect_size = base_size/22){
  # theme_light() %+replace%
  ggthemes::theme_fivethirtyeight(base_size = base_size,
                                  base_family = base_family) %+replace%
    theme(
      plot.background = element_rect(fill = "#fcedd1", colour = "#fcedd1"),
      panel.background = element_rect(fill = "#fcedd1", colour = "#fcedd1"),
      title = element_text(colour = "black"),
      axis.title = element_text(), 
      legend.background = element_rect(fill = "#fcedd1"), 
      legend.key = element_rect(fill = "#fcedd1")
    )
}

theme_set(theme_wyt(base_size = 16))

lab.y = "~log points"
lab.x = "Month"
lab.x.year = "Years since renewal"

scale_colour_discrete <- function(...) scale_color_discrete_qualitative(palette="Dark 3")
scale_fill_discrete <- function(...) scale_fill_discrete_qualitative(palette="Dark 3")

```


class: inverse, middle

## Government funds a lot of science

--

## Government can be messy

--

## Example: US fiscal policy

---

class: middle

### The last time the US Congress passed an entire federal budget before the Fiscal Year began?

--

### 1997

---

## Science funding agencies feel it 

> "...it feels strange to be writing in early November to announce preliminary funding lines for the year. **We really, really do have a full-year budget, for the first time in 22 years**" - Dr. Robin Barr, Director, Division of Extramural Activities for NIA

--

> "The uncertainty that the NIH feels reflects itself in my willingness to hire" - [David Cheresh, UCSD cancer researcher (Jan 2018)](http://www.sandiegouniontribune.com/business/biotech/sd-me-shutdown-research-20180122-story.html)

---

## What I do

1. **Funding interruptions** - delays in when projects are funded

1. NIH-funded research

1. Transaction-level data - can observe research activity at high frequency and level of detail

---

## Labs that experienced interrupted funding:

1. Spend about 50% less in the year that funding is expiring

1. Spend about 90% less in the first month after funding resumes

1. Having additional grants cushions impact on spending

1. Interruptions lead to a 1.7% decrease in publication counts for labs with less grant support

---

## Related literature

### Public

Expiring budgets lead to spikes in year-end spending that are potentially wasteful (Liebman and Mahoney 2017)

--

### Innovation

How does uncertainty affect scientists' willingness to take risks (Azoulay, Graff-Zivin, Manso 2009)?

--

### Macro

Estimating the effect of uncertainty on firm activity (Baker, Bloom, Davis 2016)

```{r my-contribution, eval = FALSE}
df = tibble(
  y = c(3, 2, 1),
  x = 1,
  words = c("Get money", "Choose projects", "Do science"))

df %>%
  ggplot(aes(x, y, label = words)) +
  geom_label(size = 10) +
  annotate("segment", x = 1, xend = 1, y = 2.9, yend = 2.1, arrow = arrow()) +
  annotate("segment", x = 1, xend = 1, y = 1.9, yend =1.1, arrow = arrow(), linetype = 2) +
  xlim(c(0, 2)) +
  annotate("label", x = 1.5, y = 1.6, label = "Distribute funds", size = 10, fill = "#fefd11") +
  theme_void() +
  theme(
      plot.background = element_rect(fill = "#fcedd1", colour = "#fcedd1"),
      panel.background = element_rect(fill = "#fcedd1", colour = "#fcedd1"))
```

---

class: middle, inverse

## Background: Scientist perspective

---

## R01 grant mechanism

R01 grant is needed to sustain an independent lab

Usually last 4-5 years

--

No-cost extension: Can be extended an additional year if you didn't spend all your money

---

## What happens when an R01 ends?

Have to renew the project

---


## Fail to renew

```{r}
scenarios = tribble(
  ~left, ~right, ~type, ~height, ~line,
  -2,    0,      "fail",     1,       1, 
   0,    2,     "fail",      1,       2, 
  -2,    0,      "cont",     2,       3, 
   0,    2,      "cont",     2,       5, 
  -2,    0,      "interr",   3,       6, 
   2,    4,    "interr",     3,       7
)

scenarios %<>% mutate(type = fct_relevel(type, c("fail", "cont", "interr")))

plot_scenario = scenarios %>% 
  ggplot() + 
  aes(alpha = type) +
  geom_errorbar(aes(group = line, x = height, ymin = left, ymax = right), size = 1.5) + 
  coord_flip()  + 
  geom_blank() + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        axis.title = element_blank(), 
        axis.text = element_blank(), 
        legend.background = element_blank()) 

plot_scenario = plot_scenario + 
  # Big Red Cross
  geom_line(data = tibble(height = c(1.45, 0.55), y = c(0, 2), type = "fail"), 
            aes(height, y), 
            colour = "red", size = 2) + 
  geom_line(data = tibble(height = c(1.45, 0.55), y = c(2, 0), type = "fail"), 
            aes(height, y), 
            colour = "red", size = 2) + 
  # interrupted line
  geom_line(data = tibble(height = c(2.5, 3.5), y = 1, type = "interr"), 
            aes(height, y),
            colour = "red", size = 2, linetype = 2) +
  scale_alpha_manual(values = c(1, 0, 0), guide = F) + 
  NULL

plot_scenario 

```

---

## Renewed and funded continuously
```{r}
plot_scenario + 
  scale_alpha_manual(values = c(0.2, 1, 0), guide = F) 
```

---

## Renewed with interrupted funding

```{r}
plot_scenario + 
  scale_alpha_manual(values = c(0.2, 0.2, 1), guide = F) 
```

---

## Renewal timeline (no hiccups)

```{r renewal-timeline}

renewline = tibble(
  y = 0.5, 
  month = c(5, 30, 60, 80), 
  event = c("Apply", " Scientific Review", "Funding Decision", "Funding Starts"),
  funding = c(F, F, T, T), 
  monthlbl = c("0", "4", "7", "8"), 
  y_month = 0.475
)

p_renewline = renewline %>% 
  ggplot(aes(month, y, label = str_wrap(event, 10))) + 
  annotate(geom = "segment", x = 0, xend = 90, y = 0.5, yend = .5) + 
  geom_label(size = 6) + 
  geom_text(aes(label = monthlbl, y = y_month), size = 7) + 
  theme(axis.title.y = element_blank(), 
        axis.text = element_blank(),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank()
        ) + 
  ylim(0.44, 0.534) + 
  annotate(geom = "text", x = 45, y = 0.46, label = "Month", 
           size = 7) + 
  labs(x = "")


p_renewline 
```

---

## Renewal timeline

```{r}
# renewline %>% 
#   ggplot(aes(month, y, label = str_wrap(event, 10))) + 
#   annotate(geom = "segment", x = 0, xend = 90, y = 0.5, yend = .5) + 
#   geom_label(size = 6) + 
#   geom_label(size = 6, data = renewline %>% filter(funding), fill = "#add8f4") + 
#   geom_text(aes(label = monthlbl, y = y_month), size = 5) + 
#   theme(axis.title.y = element_blank(), 
#         axis.text = element_blank(),
#         panel.grid.major = element_blank(), 
#         panel.grid.minor = element_blank(),
#         legend.position = "none") + 
#   ylim(0.45, 0.55) + 
#   labs(x = "Month")

p_renewline + 
    geom_label(size = 6, data = renewline %>% filter(funding), fill = "#add8f4")
  
```

---

## Continuous vs. interrupted

```{r}
plot_scenario + 
  scale_alpha_manual(values = c(0, 1, 1), guide = F) 
```
---
class: inverse, middle, center

## Background: NIH perspective

---

## Budgeting process

NIH budget is determined every fiscal year

Has to be approved by Congress

--

NIH is made up of 27 Institutes and Centers (ICs), organized by disease or body system (e.g. National Cancer Institute, National Heart, Lung, and Blood Institute)

Each IC manages its own budget

---

## Why are there interruptions?

There would be **no interruptions** if an IC knows 

1. Budget 

2. Quantity and "quality" of projects available to be funded   

---

## Interruptions arise because of uncertainty over: 

--
.pull-left[
### Budget

Congressional stalemate over federal budget
]

.pull-right[
### Quality/quantity of projects

Multiple review cycles throughout the year
]
--

Both factors are amplified by IC risk-aversion towards spending too early

---

## Why might interruptions matter?

Funds have to last longer than originally intended 

Uncertainty about whether or when funding will resume

Research may be disrupted even when funding resumes (frictions to acquiring inputs)

---

class: inverse, middle, center

# Data

---

## Transaction data from UMETRICS

Captures transactions from sponsored projects

31 universities

Coverage from 2001 to 2018 (heaviest from late 2000s onward)

---

## What we can see in UMETRICS

Total direct expenditure

Payments to vendors, subawards

Who is paid by a grant and their occupation

Do not observe how much each person is paid, but can back out total payments to labor

---

## NIH ExPorter

Publicly available data on awards made including project start and end dates, budget size

Each project has at least one Principal Investigator (PI)

---

## Two levels of data

.pull-left[
### Inputs

- UMETRICS + ExPorter
- Monthly

]

.pull-right[
### Outputs 

- ExPorter + PubMed 
- Yearly
- How do interruptions affect productivity?
]

---

## Data setup

Identify all R01s that were renewed within the same fiscal year

Stack all R01-PIs to create panel 

In a unit-time panel, unit is an R01 renewal-PI

---

class: inverse, middle

# Empirical strategy

---

## Some definitions:

- An R01 is *interrupted* if the time between expiry and renewal was greater than 30 calendar days

--

- Count number of R01-equivalent grants between 1 year before expiry and one year after renewal

- Create two subsets: PIs with one R01 and PIs with multiple R01-equivalents

---
# Event study for inputs

Project's budget could change after being renewed

--

Event study *within* budget $\implies$ Estimate expiring and renewed periods separately

---

## Baseline Specification (expiring period)

$$y_{iRt} = \sum_{j=-22}^{0} \beta_j \textbf{1}(j = t - t_{expiry})\textbf{1}(Interrupted) + \delta_{iR} + \gamma_{t - t_{expiry}} + \epsilon_{iRtm}$$

- $t-t_{expiry} = -23, -22, \dots, -1, 0$
--
- $y_{iRt}=$ total PI spending (inverse hyperbolic sine)
--
- $\delta_{iR}$ = R01-PI fixed effects
- $\gamma_{t - t_{expiry}}$ = month from expiry fixed effects
--
- $\beta_j$ = spending relative to non-interrupted projects; excluded $t-t_{expiry} = -23$
--
- $\epsilon_{iRtm}$ = error term

---

# Threats to identification

Interrupted and non-interrupted projects have differential trends in spending unrelated to being interrupted

--

Control for time since beginning of project

---

## Augmented specification

$$y_{iRt} = \sum_{j=-22}^{0} \beta_j \textbf{1}(j = t - t_{expiry})\textbf{1}(Interrupted) + \delta_{iR} + \gamma_{t - t_{expiry}} + f(t - t_{R,start}) + \epsilon_{iRtm}$$
.pull-left[
- $t-t_{expiry} = -23, -22, \dots, -1, 0$
- $y_{iRt}=$ spending (inverse hyperbolic sine)
- $\beta_j$ = spending relative to non-interrupted projects; excluded $t-t_{expiry} = -23$
- $\delta_{iR}$ = R01-PI fixed effects
- $\gamma_{t - t_{expiry}}$ = month from expiry fixed effects
- $\epsilon_{iRtm}$ = error term]

--

.pull-right[
- $t - t_{R,start}$ = time since project start
- $f(t-t_{R,start})$ = quarters since start fixed effects
]

---

class: inverse, middle, left

# Descriptives

---

## Distribution of funding gaps
```{r}
renewed_r01s = read_csv(here::here("data/renewed_r01s_balance.csv"), col_types = cols())

renewed_r01s %>% 
  ggplot(aes((days2renew_nextperiod))) + 
  geom_histogram(aes(y=..count../sum(..count..)), binwidth = 30) + 
  labs(y = "Proportion", 
       x = "Length of funding gap (calendar days)", 
       subtitle = "Binwidth = 30", 
       caption = str_wrap("Expiring R01 project periods that lasted 2-6 years and were renewed within the same fiscal year. Source: NIH ExPorter", width = 50))
```

---

## Interruptions over time

```{r}

gap_summary = renewed_r01s %>% 
  filter(expiry_fy >= 1990, 
         expiry_fy <= 2014,
         renewed_nextperiod_after_nfy <= 0,
         TRUE) %>% 
  group_by(expiry_fy) %>% 
  summarise(n = n(),
            mean_gap = mean(days2renew_nextperiod, na.rm = TRUE),
            pct_gap = mean(days2renew_nextperiod > 30, na.rm = TRUE) * 100) 

plot_pctgap = gap_summary %>% 
  ggplot(aes(expiry_fy, pct_gap)) + 
  geom_point(size = 4) + 
  geom_line() + 
  geom_vline(xintercept = 1996, linetype = 2) + 
  annotate(geom = "label", x = 1996, y = 24, label = "Shutdown", size = 5) + 
  labs(x = "Fiscal Year", y = "Percent (%)",
       # subtitle = "Sample - R01s that expired and renewed in the same FY", 
       caption = "Source: NIH ExPorter")

plot_pctgap

```


---

## Interrupted projects are similarly funded before renewal

```{r, out.width='95%', fig.asp=0.5}
renewed_r01s %<>% 
  mutate(interrupted_label = if_else(interrupted, "Interrupted", "Continuous"))

totalcost_balance = renewed_r01s %>% 
  ggplot(aes(log(total_period_cost))) + 
  geom_density(aes(fill = interrupted_label), alpha = 1/2) + 
  scale_fill_discrete_diverging(palette = "Tropic")  + 
  labs(fill = "", 
       x = "Total R01 project period funding") + 
  theme(legend.position = c(0.9, 0.75), 
        legend.direction = "vertical", 
        legend.background = element_rect(fill = "transparent"), 
        axis.title.x = element_text(size = 13))

yearlycost_balance = renewed_r01s %>% 
  ggplot(aes(log(yearly_period_cost))) + 
  geom_density(aes(fill = interrupted_label), alpha = 1/2) + 
  scale_fill_discrete_diverging(palette = "Tropic")  + 
  labs(fill = "",
       y = "",
       x = "Yearly R01 project period funding") + 
  theme(legend.position = "none", 
        axis.title.x = element_text(size = 13)) 

totalcost_balance | yearlycost_balance 

```

---

## Interrupted projects tend to have lasted longer

```{r}

projperiodlength_balance = renewed_r01s %>% 
  count(interrupted, interrupted_label, projperiod_length_round) %>% 
  group_by(interrupted, interrupted_label) %>% 
  mutate(p = n/sum(n)) %>% 
  ggplot(aes(projperiod_length_round)) + 
  geom_col(aes(y = p, fill = interrupted_label), 
                 alpha = 1/2, 
                 position = "identity") + 
  scale_fill_discrete_diverging(palette = "Tropic")  + 
  labs(fill = "", 
       x = "Length of R01 project period", 
       y = "Proportion") + 
  theme(legend.position = c(0.2, 0.95), 
        legend.direction = "vertical", 
        # legend.background = element_rect(fill = "transparent"), 
        legend.text = element_text(size = 16))

projperiodlength_balance
```

---

# Why are interrupted projects longer?

Possible answer: PIs using no-cost extensions to avoid uncertainty of waiting for a renewal decision

---

## Concern: Differential spending trends for awards of different lengths

*Especially* when an award lasts for longer than it was supposed to

--

Spending patterns of a 5-year project are different than for a "4 + 1"-year project

---

## Spending trajectory of projects

```{r}

coefs_spending_trajectory = read_csv(here("results/coefs_r01_spending_trajectory_periodfe.csv")) 

plot_spending_trajectory = coefs_spending_trajectory %>% 
  filter(projperiod_length_round %in% 3:6) %>% 
  mutate(projlen_text = 
           str_c(projperiod_length_round, "years", sep = " ")) %>% 
  ggplot(aes(months_since_period_start, estimate)) + 
  geom_line() + 
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 1/2) + 
  scale_x_continuous(breaks = c(1, 12, 24, 36, 48, 60)) + 
  theme(strip.background = element_rect(fill = "#4671d6"), 
        strip.text.x = element_text(colour = "beige"), 
        plot.subtitle = element_text(size = 13), 
        panel.grid.minor.x = element_blank()) + 
  facet_wrap(~projlen_text) + 
  labs(subtitle = "Spending relative to first month", 
       y = lab.y, 
       x = "Months into project period") 

plot_spending_trajectory

```

<!-- --- -->

<!-- ## Spending trajectory of projects -->

<!-- ```{r} -->

<!-- plot_spending_trajectory +  -->
<!--   labs(subtitle = "highlight") -->

<!-- ``` -->

---

class: middle, inverse

# Results

---

## Total spending decreases during expiry

```{r read-totalspend-coefs}

spend_lastyear = read_csv(here::here("results/coefs_expiring_diffascoefs_by_nr01s.csv"), col_types = cols()) 
spend_firstyear = read_csv(here::here("results/coefs_renewed_diffsascoefs_by_n01s.csv"), col_types= cols())

spend_lastyear %<>% rename(month = months_from_period_expiry)
spend_firstyear %<>% rename(month = months_since_period_start)

spend_all = bind_rows(spend_lastyear, spend_firstyear)  

spend_all %<>% bind_rows(tibble(estimate = 0, multiple_r01equiv = c(F, T, T, F), month = c(-23, 24, -23, 24)))
```

```{r totalpispending-eventstudy-left}

plot_totalspend_oner01_left = spend_all %>%
  filter(!multiple_r01equiv, month <= 0) %>% 
  ggplot(aes(month, estimate)) +
  geom_point(colour = "#E16A86") +
  geom_line(colour = "#E16A86") +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), 
              alpha = 1/2, 
              colour = "#E16A86", 
              fill = "#E16A86") + 
  geom_hline(yintercept = 0, linetype = 2) +
  aes(group = month > 0) + 
  labs(y = lab.y,
       x = lab.x, 
       subtitle = "One R01"
       ) +
  scale_x_continuous(breaks = c(seq(-23, 0, 4), seq(1, 24, 4), 0), 
                     limits = c(-23, 24)) + 
  annotate(geom = "label", x = -15.5, y = -2,  label = "Expiring", size = 5) +
  annotate(geom = "label", x = 10, y = -2, label = "Renewed", size = 5) +
  geom_vline(xintercept = 0.5, linetype = 5) +
  theme(legend.position = "bottom", strip.background = element_rect(fill = "grey85")) +
  guides(colour = FALSE) 

plot_totalspend_oner01_left 
  

```

---

## Spending recovers in 3-4 months

```{r totalpispending-eventstudy}

plot_totalspend_oner01 = plot_totalspend_oner01_left %+% 
  (spend_all %>% filter(!multiple_r01equiv))

plot_totalspend_oner01

```

---

# No change in spending during expiry

```{r}

plot_totalspend_multir01_left = spend_all %>%
  filter(multiple_r01equiv, month <= 0) %>% 
  ggplot(aes(month, estimate)) +
  geom_point(colour = "#00AD9A") +
  geom_line(colour = "#00AD9A") +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), 
              alpha = 1/2, 
              colour = "#00AD9A", 
              fill = "#00AD9A") + 
  geom_hline(yintercept = 0, linetype = 2) +
  aes(group = month > 0) + 
  labs(y = lab.y,
       x = lab.x, 
       subtitle = "Multiple R01s"
       ) +
  scale_x_continuous(breaks = c(seq(-23, 0, 4), seq(1, 24, 4), 0), 
                     limits = c(-23, 24)) + 
  annotate(geom = "label", x = -15.5, y = -2,  label = "Expiring", size = 5) +
  annotate(geom = "label", x = 10, y = -2, label = "Renewed", size = 5) +
  geom_vline(xintercept = 0.5, linetype = 5) +
  theme(legend.position = "bottom", strip.background = element_rect(fill = "grey85")) +
  guides(colour = FALSE) 

plot_totalspend_multir01_left
  
```

---

## After renewal, spending recovers more gradually

```{r}
plot_totalspend_multir01 = plot_totalspend_multir01_left %+%
  (spend_all %>%
  filter(multiple_r01equiv))

plot_totalspend_multir01
```

---

## Multiple R01s cushion spending impact

```{r}
plot_totalspend_multir01 + 
  geom_line(data = spend_all %>% filter(!multiple_r01equiv), 
            colour = "#E16A86", size = 1.1) + 
  labs(subtitle = "Multiple R01s + One R01")
  
```

---

## Do these patterns look different for different spending components?

Repeat for labor and vendor payments

---

## Labor spending looks similar (One R01)

```{r, read-laborspend-eventstudy}

coefs_laborspend_expiring_eventstudy = read_csv(here::here("results/spendingbycategory/coefs_labor_expiring_eventstudy_by_nr01s.csv"))
coefs_laborspend_renewed_eventstudy = read_csv(here::here("results/spendingbycategory/coefs_labor_renewed_eventstudy_by_nr01s.csv"))

coefs_laborspend_eventstudy = bind_rows(coefs_laborspend_expiring_eventstudy %>% rename(month = months_from_period_expiry),
                                        coefs_laborspend_renewed_eventstudy %>% rename(month = months_since_period_start))

```

```{r laborspend-eventstudy}

plot_totalspend_oner01 %+%
  (coefs_laborspend_eventstudy %>% 
     filter(!multiple_r01equiv))

```

---

## Labor spending looks similar (Multiple R01s)

```{r}
plot_totalspend_multir01 %+%
  (coefs_laborspend_eventstudy %>% 
     filter(multiple_r01equiv))
```


---

## Vendor spending recovers more slowly (One R01)


```{r read-vendor-coefs}

coefs_vendorspend_expiring_eventstudy = read_csv(here::here("results/spendingbycategory/coefs_vendor_expiring_eventstudy_by_nr01s.csv"), 
                                                 col_types = cols())
coefs_vendorspend_renewed_eventstudy = read_csv(here::here("results/spendingbycategory/coefs_vendor_renewed_eventstudy_by_nr01s.csv"), 
                                                col_types = cols())

coefs_vendorspend_eventstudy = bind_rows(coefs_vendorspend_expiring_eventstudy %>% rename(month = months_from_period_expiry),
                                        coefs_vendorspend_renewed_eventstudy %>% rename(month = months_since_period_start))


```


```{r}
plot_totalspend_oner01 %+%
  (coefs_vendorspend_eventstudy %>% filter(!multiple_r01equiv)) 
```

---

## Vendor spending recovers more slowly (One R01)

```{r}
plot_totalspend_oner01 %+%
  (coefs_vendorspend_eventstudy %>% filter(!multiple_r01equiv)) + 
  geom_line(data = coefs_laborspend_eventstudy %>% filter(!multiple_r01equiv)) + 
  labs(subtitle = "One R01, labor coefficients in black")
```

---

## Vendor spending (multiple R01s)

```{r}

plot_totalspend_multir01 %+%
  (coefs_vendorspend_eventstudy %>% 
     filter(multiple_r01equiv)) + 
  labs(subtitle = "Multiple R01s") 

```

---

## Vendor spending (multiple and one R01)

```{r}

plot_totalspend_multir01 %+%
  (coefs_vendorspend_eventstudy %>% 
     filter(multiple_r01equiv)) + 
  labs(subtitle = "Multiple R01s") + 
  geom_line(data = filter(coefs_vendorspend_eventstudy, !multiple_r01equiv)) + 
  labs(subtitle = "Multiple R01s, coefficients for one R01 in black")

```

---

## Number of personnel paid by PI (one R01)

```{r read-laborcount-eventstudy}
coefs_laborcount_expiring_eventstudy = 
  read_csv(here::here("results/employee_counts/coefs_expiring_eventstudy_labor.csv"),
           col_types = cols())

coefs_laborcount_renewed_eventstudy = 
  read_csv(
    here::here("results/employee_counts/coefs_renewed_eventstudy_labor.csv"), 
    col_types = cols())

coefs_laborcount_eventstudy = bind_rows(
  coefs_laborcount_expiring_eventstudy %>% rename(month = months_from_period_expiry),
  coefs_laborcount_renewed_eventstudy %>% rename(month = relative_month))

```


```{r plot-laborcount-eventstudy}
plot_laborcount_oner01 = coefs_laborcount_eventstudy %>% 
  filter(!multiple_r01equiv) %>%  
    ggplot(aes(month, estimate)) +
  geom_point(colour = "#E16A86") +
  geom_line(colour = "#E16A86") +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), 
              alpha = 1/2, 
              colour = "#E16A86", 
              fill = "#E16A86") +
  geom_hline(yintercept = 0, linetype = 2) +
  aes(group = month > 0) + 
  labs(y = lab.y,
       x = lab.x, 
       subtitle = "Multiple R01s"
       ) +
  scale_x_continuous(breaks = c(seq(-23, 0, 4), seq(1, 24, 4), 0), 
                     limits = c(-23, 24)) + 
  annotate(geom = "label", x = -15.5, y = -0.3,  label = "Expiring", size = 5) +
  annotate(geom = "label", x = 10, y = -0.3, label = "Renewed", size = 5) +
  geom_vline(xintercept = 0.5, linetype = 5) +
  theme(legend.position = "bottom", strip.background = element_rect(fill = "grey85")) +
  guides(colour = FALSE) 

plot_laborcount_oner01

```

---

## Number of personnel paid by PI (multiple R01s)

```{r}

plot_laborcount_multir01 = coefs_laborcount_eventstudy %>% 
  filter(multiple_r01equiv) %>%  
    ggplot(aes(month, estimate)) +
  geom_point(colour = "#00AD9A") +
  geom_line(colour = "#00AD9A") +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), 
              alpha = 1/2, 
              colour = "#00AD9A", 
              fill = "#00AD9A") +
  geom_hline(yintercept = 0, linetype = 2) +
  aes(group = month > 0) + 
  labs(y = lab.y,
       x = lab.x, 
       subtitle = "Multiple R01s"
       ) +
  scale_x_continuous(breaks = c(seq(-23, 0, 4), seq(1, 24, 4), 0), 
                     limits = c(-23, 24)) + 
  annotate(geom = "label", x = -15.5, y = -0.3,  label = "Expiring", size = 5) +
  annotate(geom = "label", x = 10, y = -0.3, label = "Renewed", size = 5) +
  geom_vline(xintercept = 0.5, linetype = 5) +
  theme(legend.position = "bottom", strip.background = element_rect(fill = "grey85")) +
  guides(colour = FALSE) 

plot_laborcount_multir01

```


---

class: inverse, middle

## Results: How are lab personnel affected?

---

## Questions

- How are personnel in interrupted projects affected (e.g. graduate students may have to graduate early)?

- Personnel-PI separations could deplete team-specific capital

---

## Setup

Find all employees paid by a PI from months -23 to -18

Stack employee-R01-PIs

Track from month -17 onwards

Limit to employees on projects that could be observed in the renewal period as well

Outcome variables: 

- Paid by any NIH grants
- Paid by the same PI

---

## Grad students less likely to be on any NIH grant

```{r, read-turnover-emp}
means_emp_torenewal = read_csv(here::here("results/employee_expirytorenewal/mean_emp_expiry_to_renewal_by_occupation.csv"))

means_emp_torenewal %<>% 
  mutate(multir01_label = 
           if_else(multiple_r01equiv, 
                   "Multiple R01s", 
                   "1 R01"), 
         multir01_label = 
           fct_reorder(multir01_label, multiple_r01equiv, .fun = max), 
         occ_squished = factor(occ_squished, 
                               levels = c("FACULTY", "POST GRADUATE RESEARCH", "GRADUATE STUDENT", "RESEARCH", "RESEARCH FACILITATION", "OTHERS")))

means_emp_torenewal %<>% 
  mutate(g = str_c(interrupted, relative_month > 0))

means_emp_torenewal %<>% 
  mutate(interrupted_label = if_else(
    interrupted, "Interrupted", "Continuous"
  ))

lab.anynih.y = "Paid by any NIH grant"
lab.paidbypi.y = "Paid by PI"
```


```{r plot-turnover-emp-gradstudent, out.width='95%', fig.asp=0.5}
plot_emp_gradstudent = means_emp_torenewal %>% 
  filter(occ_squished %in% c("GRADUATE STUDENT")) %>% 
  ggplot(aes(relative_month, estimate)) + 
  geom_point() + 
  geom_line(aes(group = g)) + 
  aes(colour = interrupted_label) + 
  scale_colour_discrete_diverging(palette = "Tropic") + 
  facet_wrap(vars(multir01_label), nrow = 1) + 
  geom_vline(xintercept = 0.5, linetype = 2) + 
  scale_x_continuous(breaks = c(-11, -5, 1, 6, 12, 18, 24), 
                     limits = c(-17, 24)) +
  labs(x = lab.x, 
       y = lab.anynih.y) + 
  theme(legend.position = c(0.4, 0.75), 
        legend.direction = "vertical",
        legend.title = element_blank(),
        legend.background = element_rect(fill = "transparent"),
        strip.background = element_rect(fill = "transparent"))

plot_emp_gradstudent

```

---

## Grad student less likely to stay with PI

```{r read-turnover-empbypi}
means_empbypi_torenewal = read_csv(here::here("results/employee_expirytorenewal/mean_empbypi_expiry_to_renewal_by_occupation.csv"))

means_empbypi_torenewal %<>% 
  mutate(multir01_label = 
           if_else(multiple_r01equiv, 
                   "Multiple R01s", 
                   "1 R01"), 
         multir01_label = 
           fct_reorder(multir01_label, multiple_r01equiv, .fun = max), 
         occ_squished = factor(occ_squished, 
                               levels = c("FACULTY", "POST GRADUATE RESEARCH", "GRADUATE STUDENT", "RESEARCH", "RESEARCH FACILITATION", "OTHERS")))

means_empbypi_torenewal %<>% 
  mutate(g = str_c(interrupted, relative_month > 0))

means_empbypi_torenewal %<>% 
  mutate(interrupted_label = if_else(
    interrupted, "Interrupted", "Continuous"
  ))
```


```{r plot-turnover-empbypi-gradstudent, out.width='95%', fig.asp=0.5}
plot_emp_gradstudent %+%
  (means_empbypi_torenewal %>% 
  filter(occ_squished %in% c("GRADUATE STUDENT"))) + 
  labs(y = lab.paidbypi.y)

```

---
## Post graduate researchers seem less affected

```{r, out.width='95%', fig.asp=0.5}
plot_emp_gradstudent %+%
  (means_emp_torenewal %>% 
  filter(occ_squished %in% c("POST GRADUATE RESEARCH"))) + 
  labs(y = lab.anynih.y)
```

---
## Post graduate researchers seem less affected

```{r, out.width='95%', fig.asp=0.5}
plot_emp_gradstudent %+%
  (means_empbypi_torenewal %>% 
  filter(occ_squished %in% c("POST GRADUATE RESEARCH"))) + 
  labs(y = lab.paidbypi.y)
```

---

## Non-faculty research

```{r, out.width='95%', fig.asp=0.5}
plot_emp_gradstudent %+%
  (means_empbypi_torenewal %>% 
  filter(occ_squished %in% c("RESEARCH"))) + 
  labs(y = lab.paidbypi.y)
```

---

## Faculty

```{r, out.width='95%', fig.asp=0.5}
plot_emp_gradstudent %+%
  (means_empbypi_torenewal %>% 
  filter(occ_squished %in% c("FACULTY"))) + 
  labs(y = lab.paidbypi.y)
```

---

class: inverse, middle

## Results: Research productivity

---

## Setup

Renewed R01-PI year panel, 5 years before and after renewal

Estimate a variant of the two-way FE estimator

---

## Problem with Two-way Fixed Effects

Traditional two-way FE would have R01-PI and year FE

.pull-left[
### Problem

- Uses variation across "treatment cohorts"]

.pull-right[
### Solution

- Year-by-treatment cohort FE 
- I.e. give each treatment cohort its own set of year FE
- We know the "treatment" date for the control group
]

---
## Publications

```{r, out.width='90%', fig.asp=0.5}
coefs_pubs_dd_eventstudy = 
  read_csv(here::here("results/coefs_piyearfe_meanpubs.csv"), 
           col_types = cols())

coefs_pubs_dd_eventstudy %<>% 
  mutate(multir01_label = 
           if_else(multir01, 
                   ">1 R01-equivalent", 
                   "1 R01"), 
         multir01_label = 
           fct_reorder(multir01_label, multir01, .fun = max))

plot_pubs_eventstudy = coefs_pubs_dd_eventstudy %>% 
  ggplot(aes(years_since_expiry, estimate)) + 
  geom_point() + 
  geom_line() + 
  aes(colour = multir01_label, 
      shape = multir01_label, 
      fill = multir01_label) + 
  scale_x_continuous(breaks = c(seq(-5, 5, by = 2)),
                     minor_breaks = NULL) + 
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 1/2) + 
  facet_wrap(~multir01_label) + 
  geom_hline(yintercept = 0, linetype = 2) + 
  labs(x = lab.x.year, 
       y = lab.y, 
       subtitle = "Event study") + 
  theme(legend.position = "none")

plot_pubs_eventstudy
```

---

## PIs with one R01 more affected
```{r}
coefs_pubs_dd_static = 
  read_csv(here::here("results/coefs_piyearfe_static_npubs.csv"),
           col_types = cols())

coefs_pubs_dd_static %<>% 
  mutate(multir01_label = 
           if_else(multir01, 
                   ">1 R01-equivalent", 
                   "1 R01"), 
         multir01_label = 
           fct_reorder(multir01_label, multir01, .fun = max)) 

plot_pubs_ddstatic = coefs_pubs_dd_static %>% 
  ggplot(aes(multir01_label, estimate)) + 
  geom_point(size = 3) + 
  aes(colour = multir01_label, 
      shape = multir01_label) + 
  scale_colour_discrete_qualitative() + 
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), 
                width = 1/10) + 
  geom_hline(yintercept = 0, linetype = 2) + 
  labs(x = "Number of R01-equivalents",
       y = lab.y, 
       subtitle = "Static/Pooled", 
       colour = "Number of R01-equivalents", 
       shape = "Number of R01-equivalents") + 
  theme(panel.grid.minor = element_blank(), 
        legend.position = "bottom") 

plot_pubs_ddstatic 

```

---

## How big is 1.7%?

Loss of a "superstar" collaborator leads to 5-8% decrease in quality-adjusted publiations (Azoulay, Graff-Zivin, Wang 2010)

Not an apples-to-apples comparison - AGW estimate average effect over 15 years

---

class: inverse, middle

# Discussion

---
## Delayed funding has costs

What this paper doesn't say: optimal level of funding delays

To the extent that at least some interruptions are the result of macro factors (e.g. shutdown, delayed budget due to unrelated policy disputes), current levels are unlikely to be optimal

---

## Interruptions are (to some extent) a policy choice

National Institute for Allergies and Infectious Diseases (NIAID) prioritizes funding projects in time

> *If NIAID is operating under a CR, that won’t necessarily delay your award...we are assiduous about issuing awards* using funds from the CR. For example, by December 21, 2016, NIAID had already issued more than twice as many competing fiscal year 2017 grants as any other institute.

---

## NIAID being assiduous

```{r}
renewed_r01s %<>%
  tidyr::extract(r01period_id, 
                 "ic", 
                 regex = "^R01([A-Z]{2})", 
                 remove = FALSE)

gap_by_ic = renewed_r01s %>% 
  filter(expiry_fy >= 1990, 
         expiry_fy <= 2014, 
         renewed_nextperiod_after_nfy == 0, 
         TRUE) %>% 
  group_by(expiry_fy, ic) %>% 
  summarise(n = n(),
            mean_gap = mean(days2renew_nextperiod, na.rm = TRUE),
            pct_gap = mean(days2renew_nextperiod > 30, na.rm = TRUE) * 100) 

gap_by_ic %>% 
  filter(ic %in% c("CA", "AI")) %>% 
  mutate(ic = if_else(ic == "CA", "NCI", "NIAID")) %>% 
  ggplot(aes(expiry_fy, pct_gap)) + 
  aes(colour = ic, shape = ic) + 
  geom_point(size = 4) + 
  geom_line() + 
  geom_vline(xintercept = 1996, linetype = 2) + 
  annotate(geom = "label", x = 1996, y = 22, label = "Shutdown", size = 5, 
           alpha = 0) +
  theme(legend.position = "bottom") + 
  labs(x = "Fiscal Year", y = "Percent (%)",
       # subtitle = "Sample - R01s that expired and renewed in the same FY", 
       caption = "Source: NIH ExPorter", 
       colour = "NIH Institute and Center", 
       shape = "NIH Institute and Center") + 
  scale_colour_discrete_qualitative(palette = "Cold")
```

---
# Conclusion

Understanding interaction of uncertainty and innovation is important

With high-frequency data, can observe effects of funding interruptions in "real time" and unpack how scientific production function responds

Transaction data suggests interruptions are disruptive; publication data suggest this can in turn hurt research productivity



